Depletion capabilities in the OpenMC Monte Carlo particle transport code

### 2020-06-04 manuscript.tex
Minor edits

OpenMC will setup tallies in a problem based on what transmutation reactions are
available in a depletion chain file, so any arbitrary number of transmutation
reactions can be tracked. By default, the automatic depletion chain generation
described in \cref{sec:depletion-chain} will include the following transmutation
reactions: fission, (n,$\gamma$), (n,2n), (n,3n), (n,4n), (n,p), and
(n,$\alpha$). While this is suitable for fission applications, fusion
applications may require a wider set of transmutation reactions to be tracked.

@@

(n,$\gamma$) reactions may result in a product nucleus being in either the
ground state or an excited state. If the excited state is a metastable state
(having an unusually long half-life), it is important to account for this
difference during depletion. The most well-known example is capture in
$^{241}$Am, which can produce either $^{242}$Am or $^{242m}$Am. Because the
metastable state has a significantly longer half-life than the ground state, it
is important to account for the branching of the capture reaction in $^{241}$Am
to accurately predict the production of higher actinides~\citep{haeck2012nse}.
This is complicated by the fact that the branching ratio may depend on the
incident neutron energy causing capture.

OpenMC does not currently allow energy-dependent capture branching ratios.
However, the depletion chain file does allow a reaction to be listed multiple
times with different branching ratios that result in different products.
Spectrum-averaged capture branching ratios have been computed in LWR and SFR
spectra and are available at \url{https://openmc.org/depletion-chains}. Within
the Python API, the class representing a depletion chain allows branching ratios
to be easily retrieved and changed. We note that while this treatment doesn't
accurately account for the change in the capture branching ratio with increasing
burnup (due to spectral changes), these changes are currently much smaller than
the differences in capture branching ratios between nuclear data evaluations.

@@

Like capture branching ratios, fission product yields (FPYs) also depend on the
incident neutron energy. ENDF fission product yield sublibraries typically
include yields tabulated at two or three energies (thermal, fast, and fusion
energies). Complicating matters is the fact that while these data are specified
for a single energy, they actually represent the observed FPYs for a spectrum of
energies (characterized by a single point). For example, FPYs are often given at
\SI{0.0253}{\electronvolt}, but this actually represents the observed FPY under
a thermal neutron spectrum. The ENDF-6 formats manual~\citep{trkov2018bnl}
recommends using linear-linear interpolation between energy points. However, at
the end of a transport solve, we typically have a one-group or multigroup
fission reaction rate---it's not clear how one should use this information to
calculate an effective FPY.

@@

Rather than choose a single method, OpenMC includes three methods for treating
the energy dependence of FPY. In the first method, FPY data corresponding to a
specific energy is manually selected. This approach may be appropriate if it is
known that neutrons in the system under consideration have a typical thermal,
fast, or fusion energy spectrum. In the second method, fission rates are tallied
above and below a specified cutoff energy. Then, thermal FPYs are used for all
fissions below the cutoff energy, and fast FPYs are used for all fissions above
the cutoff energy. In the third method, the average energy at which fission
events occur is tallied,
\begin{equation}
  \bar{E} = \frac{\int_0^\infty dE \; E \sigma_f(E)\phi(E)}{\int_0^\infty \sigma_f(E) \phi(E)},
\end{equation}
where $\sigma_f$ is the microscopic fission cross section and $\phi$ is the
neutron flux. An effective FPY is then calculated by linearly interpolating
between tabulated FPYs using the average energy. This approach is effectively
the same as that used by ORIGEN~\citep{gauld2011nt}.

@@

The abstract \texttt{DepSystemSolver} class represents a solver for
\cref{eq:depletion-matrix} when $\vect{A}$ is constant (a solver that can
evaluate \cref{eq:constant-A}). Subclasses of \texttt{DepSystemSolver} are
required to implement a method that takes the burnup matrix $\vect{A}$, an
initial density vector $\vect{n}_0$, and a timestep size, and returns the
resulting nuclide densities at the end of the step. The only subclass of
\texttt{DepSystemSolver} at present is \texttt{IPFCramSolver}, which implements
the IPF form of CRAM to evaluate \cref{eq:constant-A}. Because the CRAM method
requires solving multiple sparse linear systems, this class relies on the
third-party \texttt{scipy.sparse} package for performing a sparse linear solve
(REF).

@@

>>>>

OpenMC will setup tallies in a problem based on what transmutation reactions are
available in a depletion chain file, so any arbitrary number of transmutation
reactions can be tracked. By default, the automatic depletion chain generation
described in \cref{sec:depletion-chain} will include the following transmutation
reactions: fission, (n,$\gamma$), (n,2n), (n,3n), (n,4n), (n,p), and
(n,$\alpha$). While this is suitable for fission applications, fusion
applications may require a wider set of transmutation reactions to be tracked.
Currently, reaction rates are determined by directly tallying the reaction rates
using a tracklength estimator.

@@

(n,$\gamma$) reactions may result in a product nucleus being in either the
ground state or an excited state. If the excited state is a metastable state
(having an unusually long half-life), it is important to account for this
difference during depletion. The most well-known example is capture in
$^{241}$Am, which can produce either $^{242}$Am or $^{242\text{m}}$Am. Because the
metastable state has a significantly longer half-life than the ground state, it
is important to account for the branching of the capture reaction in $^{241}$Am
to accurately predict the production of higher actinides~\citep{haeck2012nse}.
This is complicated by the fact that the branching ratio may depend on the
incident neutron energy causing capture.

OpenMC does not currently implement a method to handle energy-dependent capture
branching ratios directly. However, the depletion chain file does allow a
reaction to be listed multiple times with different branching ratios that result
in different products. Spectrum-averaged capture branching ratios have been
computed in LWR and SFR spectra and are available at
\url{https://openmc.org/depletion-chains}. Within the Python API, the class
representing a depletion chain allows branching ratios to be easily retrieved
and changed. We note that while this treatment doesn't accurately account for
the change in the capture branching ratio with increasing burnup (due to
spectral changes), these changes are generally much smaller than the differences
in capture branching ratios between nuclear data evaluations.

@@

Like capture branching ratios, fission product yields (FPYs) also depend on the
incident neutron energy. ENDF fission product yield sublibraries typically
include yields tabulated at two or three energies (thermal, fast, and fusion
energies). Complicating matters is the fact that while these data are specified
for a single energy, they actually represent the observed FPYs for a spectrum of
energies (characterized by a single point). For example, FPYs are often given at
\SI{0.0253}{\electronvolt}, but this actually represents the observed FPY under
a thermal neutron spectrum that contains a continuum of energies. The ENDF-6
formats manual~\citep{trkov2018bnl} recommends using linear-linear interpolation
between energy points. However, at the end of a transport solve, we typically
have a one-group or multigroup fission reaction rate---it's not clear how one
should use this information to calculate an effective FPY.

@@

Rather than choose a single method, OpenMC includes three methods for treating
the energy dependence of FPYs. In the first method, FPY data corresponding to a
specific energy is manually selected. This approach may be appropriate if it is
known that neutrons in the system under consideration have a typical thermal,
fast, or fusion energy spectrum. In the second method, fission rates are tallied
above and below a specified cutoff energy. Then, thermal FPYs are used for all
fissions below the cutoff energy, and fast FPYs are used for all fissions above
the cutoff energy. In the third method, the average energy at which fission
events occur is tallied,
\begin{equation}
  \bar{E} = \frac{\int_0^\infty dE \; E \sigma_f(E)\phi(E)}{\int_0^\infty \sigma_f(E) \phi(E)},
\end{equation}
where $\sigma_f$ is the microscopic fission cross section and $\phi$ is the
neutron flux. An effective FPY is then calculated by linearly interpolating
between tabulated FPYs using the average energy. This approach is effectively
the same as that used by ORIGEN~\citep{gauld2011nt}.

@@

The abstract \texttt{DepSystemSolver} class represents a solver for
\cref{eq:depletion-matrix} when $\vect{A}$ is constant (a solver that can
evaluate \cref{eq:constant-A}). Subclasses of \texttt{DepSystemSolver} are
required to implement a method that takes the burnup matrix $\vect{A}$, an
initial density vector $\vect{n}_0$, and a timestep size, and returns the
resulting nuclide densities at the end of the step. The only subclass of
\texttt{DepSystemSolver} at present is \texttt{IPFCramSolver}, which implements
the IPF form of CRAM to evaluate \cref{eq:constant-A}. Because the CRAM method
requires solving multiple sparse linear systems, this class relies on the
third-party \texttt{scipy.sparse} package~\cite{virtanen2020nm} for performing a
sparse linear solve.

@@

\subsection{PWR Pincell}

\subsection{SFR Assembly}
